---
title: "Analysing overlaps"
author: "Dave"
format: html
editor: visual
---

## Libraries

```{r libraries}
library(here)
library(fs)

library(sf)
library(dplyr)
library(ggplot2)
library(stringr)
library(purrr)
library(tidyr)

library(carutools)
```

## Read in Data

### Read in CoM shapes

```{r readInData}
com_dirs <- dir_ls(here("geospatial-data", "com-areas"))
com_names <- str_extract(com_dirs, "(?<=/)[^/]+(?=$)")

names(com_dirs) <- com_names

com_shapes <- map(com_dirs, read_sf)
```

### Read in diocesan shapes

```{r readInDioceses}
#| cache: true
dioceses <- read_sf(
  here("geospatial-data", "coi-dioceses-as-small-area-conglomerations.geojson"))
```

### Reproject CoM shapes to diocesan crs

```{r reproject}
com_shapes <- map(com_shapes, \(x) st_transform(x, st_crs(dioceses)))
```

And bind_rows

```{r bindRowsCom}
com <-
  bind_rows(com_shapes) |> 
  select(-id) |> 
  mutate(com_name = com_names) |> 
  relocate(com_name) 

rm(com_shapes, com_names, com_dirs)
```

### Read in census data

Note that these populations are from different times, so this is a bit cheeky.

```{r readCensusData}
ni_pop <- readRDS(
  here("census-data", "northern-ireland_age-five-year_dz_2021.RDS"))

roi_pop <- readRDS(
  here("census-data", "republic-of-ireland_age_sa_2016.RDS")
)

ni_pop_tmp <- 
  select(ni_pop, -dz_2021_code) |> 
  as.matrix() |> 
  rowSums()

roi_pop_tmp <- 
  select(roi_pop, -sa_2015_code) |> 
  as.matrix() |> 
  rowSums()

ni_pop = tibble(code = ni_pop[["dz_2021_code"]],
                population = ni_pop_tmp,
                type = "NI data zone 2021")

roi_pop = tibble(code = roi_pop[["sa_2015_code"]],
                 population = roi_pop_tmp,
                 type = "ROI small area 2015")

population = 
  rbind(ni_pop, roi_pop) |> 
  mutate(
    type = factor(type),
    code = str_remove(code, "SA2017_"))

rm(ni_pop, ni_pop_tmp, roi_pop, roi_pop_tmp)

dir_create(here("data"))
saveRDS(population, here("data", "ni-and-roi_small-area-populations.RDS"))
```

## Plot

### Get overlapping dioceses

```{r plot}

intersections <- st_intersects(com, dioceses, sparse = FALSE)

com <- mutate(com, row_num = row_number())
com <- rowwise(com)

dioceses <- rename(dioceses, sa_geometry = geometry)

com <- 
  mutate(
    com,
    intersecting_small_areas =
      list(
        dioceses[
          intersections[row_num, ],
          c("sa_geometry", "area_code", "diocese")
          ]
        )) |>

  unnest(intersecting_small_areas) |> 
  select(-row_num)

ggplot(com) + 
  geom_sf(aes(geometry = sa_geometry, fill = diocese, colour = com_name)) +
  ca_scale_fill_discrete() + 
  ca_scale_colour_discrete()
```

Okay for the sake of simplicity let's just say that Newry belongs to Armagh Diocese. What do these dioceses look like?

```{r plotDioceses}
dioceses |> 
  filter(diocese %in% c("Armagh", "Connor")) |> 
  ggplot() +
  geom_sf(aes(colour = type, fill = diocese)) + 
  ca_scale_fill_discrete()
```

Okay so the Diocese of Armagh is partly in the Republic of Ireland. Let's find out what % of its population lives there.

## Join population data

```{r joinPopulationData}
dioceses <- 
  select(dioceses, -type) |> 
  left_join(population, by = c("area_code" = "code")) 

dioceses |> 
  as.data.frame() |> 
  select(-sa_geometry) |> 
  filter(diocese == "Armagh") |> 
  summarise(population = sum(population), .by = type) |> 
  mutate(percent = scales::percent(population/sum(population)))
```

Okay so about a quarter of the population of the diocese of Connor live in the Republic of Ireland. How close is our CoM to the border?

### CoM location

```{r comLocation}
diocesan_areas <- 
  dioceses |> 
  filter(diocese %in% c("Armagh")) |> 
  group_by(type) |> 
  summarise(sa_geometry = st_union(sa_geometry))

com |> 
  filter(com_name == "newry") |> 
  ggplot(fill = ca_gold(), col = "black") +
  geom_sf() + 
  geom_sf(data = diocesan_areas, 
          aes(fill = type), colour = NA,
          alpha = 0.35)
```

Okay so it is very much in Northern Ireland, but close the border and in a diocese where 25% of the population live in the Republic.

## News just in!

The website says the Newry is a partnership between CA and the Diocese of Down and Dromore, so let's assume my data are wrong (or the CoM is just situated mostly outside the diocese).

```{r plotAgain}
dioceses |> 
  filter(diocese %in% c("Armagh", "Connor", "Down and Dromore")) |> 
  ggplot() +
  geom_sf(aes(colour = type, fill = diocese)) + 
  ca_scale_fill_discrete() +
  geom_sf(data = filter(com, com_name == "newry"),
          fill = "black", colour = NA)
```

Brilliant! We use the diocese of Down and Dromore and nothing needs to be complicated.
